Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-07 surfaces-write-up/surface-write-up.tex
edited all but big theorem in effective pm

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$ where $\alpha_i = \frac{c_i}{k_i}$.
Then notate
\[\ell_i = \lcm_{j\ne i}(k_i).\]. 
Let
$f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that
$D_i = V(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degrees at most 
\[
	\sum_{i=0}^n \ell_i a_i
\]

@@

In this subsection, we restrict attention to the case of effective
fractional divisors $D \in \bq \otimes \di \bp^m$. We give an explicit presentation of the canonical rings when $D$ is an effective
divisor.

\begin{convention}
Let $[x_0, \ldots, x_m]$ be coordinates for $\bp^m$. Let
$\vec{v} = (v_0, \ldots, v_m) \in \bz^{m + 1}$.  Then write
\[
	x^{\vec{v}} := \prod_{i = 0}^{m} x_i^{v_i}
\]
\end{convention}

@@

\noindent
be the set of vectors of degree $c_i$.
\end{defn}

Next, we define an ordering on the vectors, which will allow us
to present a system of relations for $R_D$. Although it is not
stated explicitly later, in the case that $D = \alpha_1 D_1$, the relations in Theorem ~\ref{thm:proj-one-point} form a reduced grobner basis with respect to this ordering.

\begin{defn}
\label{defn:vec-order}
Let $\vec{v}, \vec{w} \in \bz^{m+1}$. Let $i \in \{0,\ldots, m\}$
be the biggest index such that $v_i$ is nonzero
and $j \in \{0,\ldots, m\}$ be the smallest index such that $w_j$ is
nonzero. Define $\vec{v} \prec \vec{w}$ if $i \leq j$.
\end{defn}

We are now ready to give an inductive method for computing
the generators and relations of $R_D$ in terms of $R_{D'}$, 
in the case that $D' = R_D + \alpha H$ for $H$ a hyperplane.
The statement and proofs are natural generalizations of
\cite[Theorem 6]{dorney:canonical}.

@@

The above Theorem ~\ref{thm:proj-one-point} allows us to inductively
compute presentations of effective divisors which are supported on
hyperplanes, but for $m \geq 2$, there are hypersurfaces which
are not unions of hyperplanes. We now adress this general case, giving an inductive presentation of canonical rings of effective divisors, and a tight bound on the degrees of their generators and
relations.

@@

\begin{proof}
We proceed by induction on $n$.
If $n = 0$, i.e.~ $D = 0$, then we are done.
Now, we inductively add hypersurfaces.
Let $D' = \sum_{i = 0}^{n} \alpha_i D_i \in \bq \otimes_\bq\di \bp^m$, 
for which the theorem holds. It suffices to show
the theorem holds for $D \in \di \bq \otimes_\bq \bp^m$, where $D = D' + C$ for some
degree $\pdeg$ hypersurface $C$.
If $C$ were a hyperplane, we would then be done, by Theorem ~\ref{thm:proj-one-point}.

To complete the theorem, we reduce the case that $C$ is a general hypersurface
to the case that $C$ is a hyperplane, essentially by using the Veronese embedding.

If $C$ is of degree $\pdeg$, then we construct the veronese embedding
$\nu_\pdeg^m: \bp^m \rightarrow \bp^{\binom{m+\pdeg}{\pdeg}-1}$
so that the image of $C$ is the intersection of 
a hyperplane in $\bp^{\binom{{m + \pdeg}}{\pdeg} - 1}$ with $\nu_\pdeg^m(\bp^m)$.
Now, note that the ring $H^0(\bp^m, C)$ is isomorphic to the
$\pdeg$ veronese subring of $R_{V(x_0)}$, and so
the canonical ring $R_H$ of the hyperplane $H \subset \bp^{\binom{m+\pdeg}{\pdeg}-1}$
restricting to $\nu_\pdeg^m(C)$ is isomorphic to $R_C$.
Therefore, we can bound the degree of generators and relations of
$R_D$ over $R_{D'}$ by the degree of generators and
relations for $R_C \cong R_H.$
This reduces the case of a hypersurface $C$ to
a hyperplane $H$, completeting the proof by Theorem
~\ref{thm:proj-one-point}.
\end{proof}

>>>>

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$ where $\alpha_i = \frac{c_i}{k_i}$.
Then notate
\[\ell_i = \lcm_{j\ne i}(k_i).\] 
Let
$f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that
$D_i = V(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degrees at most 
\[
	\sum_{i=0}^n \ell_i a_i
\]

@@

In this subsection, we restrict attention to the case of effective
fractional divisors $D \in \bq \otimes \di \bp^m$. We give an explicit presentation of the canonical ring when $D$ is an effective
divisor.

\begin{convention}
Let $[x_0, \ldots, x_m]$ be coordinates for $\bp^m$. Let
$\vec{v} = (v_0, \ldots, v_m) \in \bz^{m + 1}$.  Then write
\[
	x^{\vec{v}} := \prod_{i = 0}^{m} x_i^{v_i}.
\]
\end{convention}

@@

\noindent
\end{defn}

Next, we define an ordering on these vectors, which will be used to give a presentation for $R_D$. Although it is not
stated explicitly later, in the case that $D = \alpha_1 D_1$ is supported on a single hypersurface, the relations in Theorem ~\ref{thm:proj-one-point} form a reduced grobner basis with respect to this ordering.

\begin{defn}
\label{defn:vec-order}
Let $\vec{v}, \vec{w} \in \bz^{m+1}$. Let $i \in \{0,\ldots, m\}$
be the biggest index such that $v_i$ is nonzero
and $j \in \{0,\ldots, m\}$ be the smallest index such that $w_j$ is
nonzero. Define a partial ordering on $\bz^{m+1}$ by $\vec{v} \prec \vec{w}$ if $i \leq j$.
\end{defn}

We are now ready to give an inductive method for computing
the generators and relations of $R_D$ in terms of $R_{D'}$ 
in the case that $D' = R_D + \alpha H$ for $H$ a hyperplane.
The statement and proofs are natural generalizations of
\cite[Theorem 6]{dorney:canonical}.

@@

Theorem ~\ref{thm:proj-one-point} gives an inductive procedure to
compute presentations of effective divisors which are supported on
hyperplanes. However, for $m \geq 2$, there are hypersurfaces which
are not unions of hyperplanes. We now adress this general case, giving an inductive presentation of canonical rings of effective divisors, and a tight bound on the degrees of their generators and
relations.

@@

\begin{proof}
We proceed by induction on $n$.
If $n = 0$, i.e.~ $D = 0$, then we are done.
Now, we inductively add hypersurfaces.
Let $D' = \sum_{i = 0}^{n} \alpha_i D_i \in \bq \otimes_\bq\di \bp^m$, 
and assume the theorem holds for $D'$. It suffices to show
the theorem holds for $D \in \di \bq \otimes_\bq \bp^m$, where $D = D' + \alpha C$ for some
degree $\pdeg$ hypersurface $C$.
If $C$ were a hyperplane, we would then be done, by Theorem ~\ref{thm:proj-one-point}.

To complete the theorem, we reduce the case that $C$ is a general hypersurface
to the case that $C$ is a hyperplane, by using the Veronese embedding.

If $C$ is of degree $\pdeg$, consider the veronese embedding
$\nu_\pdeg^m: \bp^m \rightarrow \bp^{\binom{m+\pdeg}{\pdeg}-1}$
so that the image of $C$ is the intersection of 
a hyperplane in $\bp^{\binom{{m + \pdeg}}{\pdeg} - 1}$ with $\nu_\pdeg^m(\bp^m)$.
Now, note that the ring $R_{\alpha C}$ is isomorphic to the
$\pdeg$ veronese subring of $R_{\alpha V(x_0)} = \bigoplus_{d \geq 0} H^0(\bp^m, d \alpha V(x_0))$ and so
the canonical ring $R_H$ of the hyperplane $H \subset \bp^{\binom{m+\pdeg}{\pdeg}-1}$
restricted to $\nu_\pdeg^m(C)$ is isomorphic to $R_C$.
Therefore, we can bound the degree of generators and relations of
$R_D$ over $R_{D'}$ by the degree of generators and
relations for $R_C \cong R_H.$
This reduces the case of a hypersurface $C$ to
a hyperplane $H$, completeting the proof by Theorem
~\ref{thm:proj-one-point}.
\end{proof}
